\pkg{dataMaid}: your assistant for documenting supervised data quality screening in \proglang{R}

### 2017-11-08 latex/article_vol3.Rnw
First round of tests

%% almost as usual
\author{Anne Helby Petersen\\Section of Biostatistics\\Department of Public
  Health\\University of Copenhagen \And Claus Thorn Ekstr\o m\\Section of Biostatistics\\Department of Public
  Health\\University of Copenhagen}
\title{\pkg{dataMaid}: your assistant for data documentation in \proglang{R}}

@@

  We describe an \proglang{R} package, \pkg{dataMaid}, which
  implements an extensive and customizeable suite of quality
  assessment tools that can be applied to a dataset in order to
  identify potential problems in its variables. The results are 
  presented in an auto-generated, non-technical, stand-alone overview
  document intended to be perused by an investigator with an
  understanding of the variables in the data, but not necessarily
  knowledge of \proglang{R}. Thereby, \pkg{dataMaid} aids the dialogue
  between data analysts and field experts, while also providing easy
  documentation of reproducible data cleaning steps and data quality
  control. Moreover, the \pkg{dataMaid} solution changes the data
  cleaning process from the usual ad hoc approach to a systematic,
  well-documented endeavor.  \pkg{dataMaid} also provides a suite of
  more typical \proglang{R} tools for interactive data quality
  assessment and -cleaning, where the data inspections are executed directly 
  in the \proglang{R} console.

  % The \pkg{dataMaid} package is designed to be easily extended with
  % custom user-created checks that are relevant in particular
  % situations. \hl{Already said that above. Either expand on it or
  % delete this.}
}
\Keywords{data cleaning, quality control, \proglang{R}, data documentation}
\Plainkeywords{data cleaning, quality control, R, data documentation} %% without formatting
%% at least one keyword must be supplied

@@

\section{Introduction}
Though data cleaning might be regarded as a somewhat tedious activity,
adequate data cleaning is crucial in any data analysis. With
ever-growing dataset sizes and complexities, statisticians and data
analysts find themselves spending a large portion of their time on
data cleaning and data wrangling. While a computer should generally not
make unsupervised decisions on what should be done to potential
errors in a dataset, it can still be an extremely useful tool in the
data cleaning process. Some errors can be tracked down and flagged by a
computer without further ado, while other types of errors need an analytic
context in order to be identified. Even in this latter case, well-designed
software can aid the process tremendously by providing the human
investigator with the information needed for identifying issues.

But even when tools are available for identifying problems in a dataset, the activity of data cleaning still suffers from a challenge that has recieved increasing attention in the scientitic communities in the later years: Data cleaning is not very straight forward to document and therefore, reproducibility suffers. We present a new \proglang{R} package, \pkg{dataMaid} \citep{dataMaid}, whose most central purpose is to facilitate a data cleaning workflow where documentation is thoroughly integrated rather than a secondary add-on. This is accomplished by centering the data cleaning process around auto-generated data overview reports that summarize the contents of the current version of a dataset and flags potential problems that could be candidates for the next bullet on the cleaning agenda.

A number of \proglang{R} packages made for other pre-analysis steps are
already available, including \pkg{janitor} \citep{janitor}, \hl{assertive}, \pkg{dplyr}
\citep{dplyr}, \pkg{tidyr}\citep{tidyr}, \pkg{data.table} \citep{data.table}, \pkg{DataCombine}
\citep{DataCombine} , \pkg{validate} \citep{validate}, \pkg{assertr} \citep{assertr}, and
\pkg{DataExplorer} \citep{DataExplorer}. These packages focus on
different stages of the pre-analysis work, most of which are not
really data cleaning.  \pkg{janitor} provides tools for data import
with a particular emphasis on the challenges of getting neat data
frames from Microsoft Excel data files. \hl{Mention assertive here}. \pkg{dplyr}, \pkg{tidyr}, \pkg{data.table}
and \pkg{DataCombine} go a few steps further by providing a wide array
of extremely powerful tools for data wrangling, including a number of
particularly useful functions for merging and working with very large
datasets. When it comes to actual data cleaning, however, the options
are fewer.  \pkg{validate} (and the similar packages \pkg{editrules}
\citep{editrules} and \pkg{deducorrect} \citep{deducorrect} from the
same authors) does offer a few tools for identifying and
auto-correcting errors in a dataset, but, as the name suggests, the
focus is on internal validity rather than general data cleaning. In
practice, this means that quite elegant tools for, e.g., linear
restraints among the variables in a dataset can be applied, but
looking for potentially miscoded missing values is not really
feasible. The main difference between these two challenges is the
direction in which the data is inspected: While linear constraints
work observation-wise with no ambiguity, determining whether or not something
is a miscoded missing value often requires knowledge about the full
variable (e.g. range or data type), and thus it should be performed
variable-wise. \pkg{validate} does not allow for user-defined extensions
of the latter type, thereby limiting its data cleaning potential.
Automatic data correction functions are also provided by
\pkg{validate} which we consider to be quite a dangerous cocktail: all
power is given to the the computer with no human supervision, and
investigators are less likely to make an active, case-specific choice
regarding the handling of the potential errors.  Finally, no tools
have been made to easily document exactly which checks and preliminary
results were used in the data cleaning process. The \pkg{assertr} package provide very similar - and very nice - tools to those of \pkg{validate}, but without any amibitions of conducting auto-cleaning.

@@

But no matter how clever software tools we make, data cleaning remains
to be a time consuming endeavor, as it inherently requires human
interaction since every dataset is different and the variables in the
dataset can only be understood in the proper context of their
origin. This often requires a collaborative effort between an expert
in the field and a statistician or data scientist, which may be why
the process of proper data cleaning is not always undertaken. In many
situations, these errors are discovered in the process of the data
analysis (e.g., a categorical variable with numeric labels for each
category may be wrongly classified as a quantitative variable or a
variable where all values have erroneously been coded to the same
value), but in other cases a human with knowledge about the data
context area is needed to identify possible mistakes in the data
(e.g., if there are 4 categories for a variable that should only have
3).

The \pkg{dataMaid} approach to data cleaning, -quality assessment and -documentation is
governed by two fundamental paradigms. First of all, there is no need
for data cleaning to be an ad hoc procedure. Often, we have a very
clear idea of what flags are raisable in a given dataset before we
look at it, as we were the ones to produce it in the first place. This
means that data cleaning can easily be a well-documented,
well-specified procedure. In order to aid this paradigm, \pkg{dataMaid}
provides easy-to-use, automated tools for data quality assessment in
\proglang{R} \citep{R} on which data cleaning decisions can be made. This quality
assessment is presented in an auto-generated overview document,
readable by data analysts and field experts alike, thereby also
contributing to an inter-field dialogue about the data at
hand. Oftentimes, e.g., distinguishing between faulty codings of a
numeric value and unusual, but correct, values requires
problem-specific expertise that might not be held by the data
analyst. Hopefully, having easy access to data descriptions through
\pkg{dataMaid} will help this necessary knowledge sharing.

This manuscript is structured as follows: First, in Section
\ref{sec:usingdataMaid}, we introduce the main representative of the first
paradigm, namely the \code{makeDataReport()} function, which generates data
overview documents. In the \pkg{dataMaid} package, we have
provided a number of default checks that cover the data
cleaning challenges we find to be most common and these are also summarized in Section \ref{sec:usingdataMaid}. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{dataMaid}, as motivated
by the second paradigm above. Next, we show atep-by-step how the data report and the interactive mode of \pkg{dataMaid} can be combined to conduct a well-documented, systematic data cleaning in Section \ref{sec:bigExample}. Here, we assess and clean a dirty dataset with information about the US presidents.  At last, in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning- and documentation challenges and how \pkg{dataMaid} can be used to
solve them.

Please note that this description of what \pkg{dataMaid} contains is usefully supplemented by a tutorial into how it can be extended. Every dataset is different, and some datasets might include problems that cannot be detected by our data checking functions. Therefore, \pkg{dataMaid} was designed to be easily extendend with user-supplied functions for summarizing, viusalizing and checking data. We have provided a vignette \hl{REF TO VIGNETTE} in which we describe how \pkg{dataMaid} extensions can be made, such that they are integrable with the \code{makeDataReport()}
function and with the other tools available in \pkg{dataMaid}.

@@

The \code{makeDataReport()} function is the primary workhorse of \pkg{dataMaid} and
this is the only function needed if a standard battery
of tests are used to generate data reports. The data report itself is an overview document, intended for reading
by humans, in either pdf, html or word (.docx) format. Appendix \ref{sec:appendix1}
provides an example of a data report, produced by
calling \code{makeDataReport()} on the dataset \code{toyData} available in
\pkg{dataMaid}. The first two pages (excluding the frontpage) of this data report are
shown in Figure~\ref{fig:example1} and the following two pages are shown in Figure~\ref{fig:example2}. \code{toyData} is a very
small ($15$ observations of $6$ variables), artificial dataset which was created to
illustrate the main capabilities of \pkg{dataMaid}. The following
commands load the dataset and produce the report:

@@

By default, an \proglang{R} markdown file and a rendered pdf \hl{OBS: måske html/doc alt efter hvem der kalder funktionen?} overview
document is produced, saved to the working directory and
opened for immediate inspection. Such a data report consists of three parts, two of which 
are presented in Figure~\ref{fig:example1}. First, an overview of what was done is presented under the
title \textit{Data report overview}. Secondly, an index listing each variable along with an indication of whether it was found to be problematic or not is provided. Thirdly, as seen in Figure~\ref{fig:example2}, each variable in the
dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution and a
list of flagged issues, if any. For instance, as shown in Figure~\ref{fig:example2}, for the \code{numeric}-type variable
\code{var2} from \code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and \code{NaN}),
while two values were also flagged as potential outliers that should
be investigated more carefully.

Though the \code{makeDataReport()} function is very easy to use, it should not be
mistaken to be inflexible: By using the optional arguments of the function, 
both the contents and the look of the data report can be molded 
according to the user's needs. The most commonly used arguments are summarized in
Table~\ref{table.cleanFormals} and they are grouped according to the
part of the data assesment and report generation they influence. In order to understand
this distinction, a glimpse of the inner structure of \code{makeDataReport()} is
shown in Figure~\ref{figure:cleanStructure}. Below, we present a few
examples on how to use the arguments from Table \ref{table.cleanFormals}
 to influence the output of a \code{makeDataReport()} call.

@@

\subsection{Polishing off the arguments}
We begin with an example that is intended as an illustration of how
\code{makeDataReport()} might be used in the very first stages of data
cleaning, when we are uncertain about the complexities of the
errors and how much time should be allocated to data cleaning. At
this stage, what is really needed, is a very rough idea of the
severity of errors in the dataset. In this scenario, we might wish to
obtain a summary document in html format that only contains the
variables with potential problems, and with a limit of, say, maximum
2 printed potential problematic value per check for each variable. Also, we can add the
argument \code{replace = TRUE} in order to force \code{makeDataReport()} to
overwrite any existing files produced by \code{makeDataReport()}.  Using the
\code{toyData} dataset as a guinea pig, we type:

@@

\subsection{Controlling contents: What summaries, visualizations and checks should be performed?}
We will now move on to discuss how not only the structure of the
data assesment steps is manipulated, but also its very contents. This is
done through the summarize/visualize/check (SVC) steps, as illustrated
in Figure \ref{figure:cleanStructure}.  \pkg{dataMaid} uses three
different types of functions for performing these steps, namely
\code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.  By default, \code{makeDataReport()} runs selected summary,
visualization and check functions on each variable in the dataset, and
the exact choice of these functions depends on the classes of the
variables. For instance, though detection of outlier values might be
interesting for numerical variables, it holds little meaning for
factor variables, and therefore, numerical and factor variables need
different checks. Table~\ref{table.SVCfunctions} lists all available
summarize/visualize/check functions, but we can also use the
\code{allSummaryFunctions()}, \code{allVisualFunctions()}, and
\code{allCheckFunctions()} functions in \pkg{dataMaid} to print overview
lists in \proglang{R}. For example, the implemented
\code{summaryFunction}s are:

@@

\subsection{Data cleaning by hand: An example}
Assume that we wish to look further into a certain variable from
\code{toyData}, namely \code{var2}. The data cleaning summary found some
issues in this variable, and we would like to recall what these issues
were. This can be done using the \code{check()} command

<<>>=
check(toyData$var2)
@

@@

<<>>=
check(toyData$var2, checks = setChecks(numeric = "identifyMissing"))
@

@@

<<>>=
identifyMissing(toyData$var2)
@

@@

<<>>=
missVar2 <- identifyMissing(toyData$var2)
str(missVar2)
@

The most important thing to note here is that while the printed
message is made for easy reading, the actual values of the variable
causing the issue are still obtainable in the entry
\code{problemValues}. If we decide that the values \code{999}
and \code{NaN} in \code{var2} are in fact miscoded missing values, we can
easily replace them with \code{NA}s:

<<>>=
toyData$var2[toyData$var2 %in% missVar2$problemValues] <- NA
identifyMissing(toyData$var2)
@

Similarly, the \code{visualize()} and \code{summarize()} functions can be
used to run the corresponding visualizations and summaries for each
variable. See Figure~\ref{fig:example2} for the visualization output.

<<>>=
visualize(toyData$var2)
summarize(toyData$var2)
@

\begin{figure}[tb]
\begin{center}
\includegraphics[width=7.5cm]{toyData-var2.pdf}
\end{center}
\caption{Output from running \code{visualize()} on the variable \code{var2} from the
\code{toyData} dataset.}
\label{fig:example2}
\end{figure}

@@

<<>>=
summarize(toyData$var2, 
  summaries = setSummaries(
  numeric = defaultNumericSummaries(remove = c("variableType", 
  "countMissing"))))
@

The syntax in this code chunk can be read as follows: "Summarize \code{var2} in \code{toyData}, and for \code{numeric} variables, set the summaries to be the default summary functions, except \code{variableType} and \code{countMissing}."

@@

Figure \ref{fig:bigExampleP45} presents the remaining two pages with variable presentations. On pages 4 and 5, we find a few remarks:
\begin{itemize}
\item In the birthday variable, there is an entry with the date March 1, 1300 which is a bit of an outlier. 
\item Among the states in which the presidents were born, New York was mistakenly spelled with a lower case "Y" in at least one entry.
\item The variable concerning assassination attempts is coded as a numeric, but the default \code{smartNum = TRUE} setting of \code{makeDataReport()} implies that such a numeric variable with only a few (less that 5) unique values is treated as a factor variable in the data report, thereby providing more relevant summaries, visualizations and checks. This is remarked in the variable presentation for \code{assassinationAttempt} and it can also be seen by the visualization being a barplot rather than a histogram.
\item The variable concerning the sex of the president was skipped, as there is nothing to present when all US presidents have been male so far.
\item The report flags the variable \code{ethnicity} to have suspiciously few observations in one category, "African American". 
\item A few presidents were found to have odd values in the variable describing the duration of their presidencies: Some had very short (outlier) presidencies of less than two years and one was registered to have an \textit{infinite} presidency. 
\item The variable concerning age at inauguration was coded as a character variable, but consists exclusively of numbers and takes a lot of different values and therefore, it was flagged as a potentially misclassified numeric variable.
\item It seems as if one or more presidents have complex numbers as their favorite numbers. As \code{complex} is not a supported variable type in \code{dataMaid} and no strategy for handling this class was provided in the \code{treatXasY} argument, the variable is simply flagged as non-supported.
\end{itemize}
A lot of these mistakes are easily fixable, and we will do so below. However, some of them require more delicate knowledge of the subject matter. For instance, \code{ethnicity} is very reasonably marked as a potentially problematic variable as it includes only a single observation of "African American". However, a human reading this report will know that this does \textit{not} reflect a mistake in the data, but rather a perculiarity in the real world, and as such, it should not be cleaned out.

@@

and we find that only a single variable in \code{toyData}, \code{var6} (for which all
observations have the value \code{"Irrelevant"}), is problem-free when using the default checks.

>>>>

%% almost as usual
\author{Anne Helby Petersen\\Section of Biostatistics\\Department of Public
  Health\\University of Copenhagen \And Claus Thorn Ekstr\o m\\Section of Biostatistics\\Department of Public
  Health\\University of Copenhagen}
\title{\pkg{dataMaid}: your assistant for documenting supervised data quality screening in \proglang{R}}

@@

  We describe an \proglang{R} package, \pkg{dataMaid}, which
  implements an extensive and customizeable suite of quality
  assessment aids that can be applied to a dataset in order to
  identify potential problems in its variables. The results are
  presented in an auto-generated, non-technical, stand-alone overview
  document intended to be perused by an investigator with an
  understanding of the variables in the data, but not necessarily
  knowledge of \proglang{R}. Thereby, \pkg{dataMaid} aids the dialogue
  between data analysts and field experts, while also providing easy
  documentation of reproducible data quality screening. Moreover, the
  \pkg{dataMaid} solution changes the data screening process from the
  usual ad hoc approach to a systematic, well-documented endeavor.
  \pkg{dataMaid} also provides a suite of more typical \proglang{R}
  tools for interactive data quality assessment and -screening, where
  the data inspections are executed directly in the \proglang{R}
  console.

  % The \pkg{dataMaid} package is designed to be easily extended with
  % custom user-created checks that are relevant in particular
  % situations. \hl{Already said that above. Either expand on it or
  % delete this.}
}
\Keywords{data screening, data cleaning, quality control, \proglang{R}, data documentation}
\Plainkeywords{data screening, data cleaning, quality control, R, data documentation} %% without formatting
%% at least one keyword must be supplied

@@

\section{Introduction}
Though data cleaning might be regarded as a somewhat tedious activity,
adequate data cleaning is crucial in any data analysis. With
ever-growing dataset sizes and complexities, statisticians and data
analysts find themselves spending a large portion of their time on
data cleaning and data wrangling. While a computer should generally not
make unsupervised decisions on what should be done to potential
errors in a dataset, it can still be an extremely useful tool in the
data cleaning process. Some errors can be tracked down and flagged by a
computer without further ado, while other types of errors need a subject
context in order to be identified. Even in this latter case, well-designed
software can aid the process tremendously by providing the human
investigator with the information needed for identifying issues.

But even when tools are available for identifying problems in a
dataset, the activity of data cleaning still suffers from a challenge
that has recieved increasing attention in the scientitic communities
in the later years: Data cleaning is not very straight forward to
document and therefore, reproducibility suffers. We present a new
\proglang{R} package, \pkg{dataMaid} \citep{dataMaid}, whose most
central purpose is to facilitate a supervised data quality screening
workflow where documentation is thoroughly integrated rather than an
add-on. This is accomplished by structuring the data screening around
auto-generated data overview reports that summarize and flags
potential problems in the dataset.

But no matter how clever software tools we make, data cleaning remains
to be a time consuming endeavor, which inherently requires human
interaction since every dataset is different and the variables in the
dataset can only be understood in the proper context of their
origin. This often requires a collaborative effort between an expert
in the field and a statistician or data scientist. In many
situations, these errors are discovered in the process of the data
analysis (e.g., a categorical variable with numeric labels for each
category may be wrongly classified as a quantitative variable or a
variable where all values have erroneously been coded to the same
value), but in other cases a human with knowledge about the data
context area is needed to identify possible mistakes in the data
(e.g., if there are 4 categories for a variable that should only have
3).

The \pkg{dataMaid} approach to data screening, -quality assessment and
-documentation is governed by two fundamental paradigms. First of all,
there is no need for data cleaning to be an ad hoc procedure. Often,
we have a very clear idea of what flags are raisable in a given
dataset before we look at it, as we were the ones to produce it in the
first place. This means that data cleaning can easily be a
well-documented, well-specified procedure. In order to aid this
paradigm, \pkg{dataMaid} provides easy-to-use, automated tools for
data quality assessment in \proglang{R} \citep{R} on which data
cleaning decisions can be made. This quality assessment is presented
in an auto-generated overview document, readable by data analysts and
field experts alike, thereby also contributing to an inter-field
dialogue about the data at hand. Oftentimes, e.g., distinguishing
between faulty codings of a numeric value and unusual, but correct,
values requires problem-specific expertise that might not be held by
the data analyst. Hopefully, having easy access to data summaries
through \pkg{dataMaid} will help this necessary knowledge sharing.

A number of \proglang{R} packages made for other pre-analysis steps
are already available, including \pkg{janitor} \citep{janitor},
\pkg{assertive} \citep{assertive}, \pkg{dplyr} \citep{dplyr},
\pkg{tidyr}\citep{tidyr}, \pkg{data.table} \citep{data.table},
\pkg{DataCombine} \citep{DataCombine}, \pkg{validate}
\citep{validate}, and \pkg{assertr} \citep{assertr}. These packages
focus on different stages of the pre-analysis work.  \pkg{janitor}
provides tools for data import with a particular emphasis on the
challenges of getting neat data frames from Microsoft Excel data
files. \pkg{dplyr}, \pkg{tidyr}, \pkg{data.table} and
\pkg{DataCombine} go a few steps further by providing a wide array of
extremely powerful tools for data wrangling, including a number of
particularly useful functions for merging and working with very large
datasets. When it comes to actual data cleaning, however, the options
are fewer.  \pkg{validate} (and the similar packages \pkg{editrules}
\citep{editrules} and \pkg{deducorrect} \citep{deducorrect} from the
same authors) and \pkg{assertive} offers tools for identifying errors
in a dataset by checking the state of the variable given a set of
pre-specified rules, and their focus is on internal validity rather
than general data screening. In practice, this means that quite elegant
tools for, e.g., linear restraints among the variables in a dataset
can be applied, but looking for potentially miscoded missing values is
not really feasible. The main difference between these two challenges
is the direction in which the data is inspected: While linear
constraints work observation-wise with no ambiguity, determining
whether or not something is a miscoded missing value often requires
knowledge about the full variable (e.g. range or data type), and thus
it should be performed variable-wise. \pkg{validate} does not
currently allow for user-defined extensions of the latter type,
thereby limiting its data cleaning potential.  Automatic data
correction functions are also provided by \pkg{validate} which we
consider to be quite a dangerous cocktail: all power is given to the
the computer with no human supervision, and investigators are less
likely to make an active, case-specific choice regarding the handling
of the potential errors.  Finally, no tools have been made to easily
document exactly which checks and preliminary results were used in the
data cleaning process. The \pkg{assertr} package provide very similar
--- and very nice --- tools to those of \pkg{validate}, but without
any amibitions of conducting auto-cleaning.

@@

This manuscript is structured as follows: First, in Section
\ref{sec:usingdataMaid}, we introduce the main representative of the first
paradigm, namely the \code{makeDataReport()} function, which generates data
overview documents. In the \pkg{dataMaid} package, we have
provided a number of default generic checks that cover the data
cleaning challenges we find to be most common and these are also summarized in Section \ref{sec:usingdataMaid}. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{dataMaid}, as motivated
by the second paradigm above. Next, we show step-by-step how the data report and the interactive mode of \pkg{dataMaid} can be combined to conduct a well-documented, systematic data cleaning in Section \ref{sec:bigExample}. Here, we assess and clean a dirty dataset with information about the US presidents.  At last, in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning- and documentation challenges and how \pkg{dataMaid} can be used to
solve them.

%Please note that this description of what \pkg{dataMaid} contains is
%usefully supplemented by a tutorial into how it can be extended. Every
%dataset is different, and some datasets might include problems that
%cannot be detected by our data checking functions. 
%Therefore,
\pkg{dataMaid} was designed to be easily extendend with user-supplied
functions for summarizing, viusalizing and checking data. We have
provided a vignette in which we describe how \pkg{dataMaid} extensions
can be made, such that they are integrate with the
\code{makeDataReport()} function and with the other tools available in
\pkg{dataMaid}.

@@

The \code{makeDataReport()} function is the primary workhorse of
\pkg{dataMaid} and it is the only function needed to generate a data
report using the standard battery of tests. The data report itself is
an overview document, intended for reading by humans, in either pdf,
html or word (.docx) format. Appendix \ref{sec:appendix1} provides an
example of a data report, produced by calling \code{makeDataReport()}
on the dataset \code{toyData} available in \pkg{dataMaid}. The first
two pages (excluding the frontpage) of this data report are shown in
Figure~\ref{fig:example1} and the following two pages are shown in
Figure~\ref{fig:example2}. \code{toyData} is a very small ($15$
observations of $6$ variables), artificial dataset which was created
with a lot of potential errors to illustrate the main capabilities of
\pkg{dataMaid}. Section~\ref{sec:bigExample} shows an example of a data sreening
process with a real dataset. The following commands load the dataset
and produce the report:

@@

By default, an \proglang{R} markdown file and a rendered pdf or word
overview document is produced, saved to the working directory and
opened for immediate inspection. Such a data report consists of three
parts, two of which are presented in Figure~\ref{fig:example1}. First,
an overview of what was done is presented under the title \textit{Data
  report overview}. Secondly, an index listing each variable along
with an indication of whether it was found to be problematic or not is
provided. Thirdly, as seen in Figure~\ref{fig:example2}, each variable
in the dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution when relevant and a list of flagged
issues, if any. For instance, as shown in Figure~\ref{fig:example2},
for the \code{numeric}-type variable \code{events} from
\code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and
\code{NaN}), while two values were also flagged as potential outliers
that should be investigated more carefully.

The arguments to \code{makeDataReport()} can be used to modify the
contents and the look of the data report according to the user's
needs. The most commonly used arguments are summarized in
Table~\ref{table.cleanFormals} and they are grouped according to the
part of the data assesment and report generation they influence. In
order to understand this distinction, a glimpse of the inner structure
of \code{makeDataReport()} is shown in
Figure~\ref{figure:cleanStructure}. Below, we present a few examples
on how to use the arguments from Table \ref{table.cleanFormals} to
influence the output of a \code{makeDataReport()} call.

@@

\subsection{Dusting off the arguments}
We begin with an example that is intended as an illustration of how
\code{makeDataReport()} might be used in the very first stages of data
cleaning, when we are uncertain about the complexities of the
errors and how much time should be allocated to data cleaning. At
this stage, what is really needed, is a very rough idea of the
severity of errors in the dataset. In this scenario, we might wish to
obtain a summary document in html format that only contains the
variables with potential problems, and with a limit of, say, maximum
2 printed potential problematic values per check for each variable. Also, we can add the
argument \code{replace = TRUE} in order to force \code{makeDataReport()} to
overwrite any existing files produced by \code{makeDataReport()}.  Using the
\code{toyData} dataset as a guinea pig, we type:

@@

\subsection{Controlling contents through summaries, visualizations and checks}

\pkg{dataMaid} works through three different steps --- summarize,
visualize, and check (SVC) --- for each variable in the dataset
(illustrated in Figure \ref{figure:cleanStructure}).  Three different
types of functions are used to perform these steps, namely
\code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.  By default, \code{makeDataReport()} runs
selected summary, visualization and check functions on each variable
in the dataset, and the exact choice of these functions depends on the
classes of the variables. For instance, detection of outlier values
might be interesting for numerical variables, but it holds little
meaning for factor variables, and therefore, numerical and factor
variables need different checks. Table~\ref{table.SVCfunctions} lists
all available summarize/visualize/check functions, but we can also use
the \code{allSummaryFunctions()}, \code{allVisualFunctions()}, and
\code{allCheckFunctions()} functions in \pkg{dataMaid} to print
overview lists in \proglang{R}. For example, the implemented
\code{summaryFunction}s are:

@@

\subsection{Data cleaning by hand: An example}
Assume that we wish to look further into a certain variable from
\code{toyData}, namely \code{events}. The data cleaning summary found some
issues in this variable, and we would like to recall what these issues
were. This can be done using the \code{check()} command

<<>>=
check(toyData$events)
@

@@

<<>>=
check(toyData$events, checks = setChecks(numeric = "identifyMissing"))
@

@@

<<>>=
identifyMissing(toyData$events)
@

@@

<<>>=
missEvents <- identifyMissing(toyData$events)
str(missEvents)
@

The most important thing to note here is that while the printed
message is made for easy reading, the actual values of the variable
causing the issue are still obtainable in the entry
\code{problemValues}. If we decide that the values \code{999}
and \code{NaN} in \code{events} are in fact miscoded missing values, we can
easily replace them with \code{NA}s:

<<>>=
toyData$events[toyData$events %in% missEvents$problemValues] <- NA
identifyMissing(toyData$events)
@

Similarly, the \code{visualize()} and \code{summarize()} functions can be
used to run the corresponding visualizations and summaries for each
variable. See Figure~\ref{fig:example3} for the visualization output.

<<toydata-events>>=
visualize(toyData$events)
summarize(toyData$events)
@

\begin{figure}[tb]
\begin{center}
\includegraphics[width=7.5cm]{toyData-events.pdf}
\end{center}
\caption{Output from running \code{visualize()} on the variable \code{events} from the
\code{toyData} dataset.}
\label{fig:example3}
\end{figure}

@@

<<>>=
summarize(toyData$events, 
  summaries = setSummaries(
  numeric = defaultNumericSummaries(remove = c("variableType", 
  "countMissing"))))
@

The syntax in this code chunk can be read as follows: "Summarize \code{events} in \code{toyData}, and for \code{numeric} variables, set the summaries to be the default summary functions, except \code{variableType} and \code{countMissing}."

@@

Figure \ref{fig:bigExampleP45} presents the remaining two pages with variable presentations. On pages 4 and 5, we find a few remarks:
\begin{itemize}
\item In the birthday variable, there is an entry with the date March 1, 1300 which is a bit of an outlier. 
\item Among the states in which the presidents were born, New York was mistakenly spelled with a lower case "Y" in at least one entry.
\item The variable concerning assassination attempts is coded as a numeric, but the default \code{smartNum = TRUE} setting of \code{makeDataReport()} implies that such a numeric variable with only a few (less that 5) unique values is treated as a factor variable in the data report, thereby providing more relevant summaries, visualizations and checks. This is remarked in the variable presentation for \code{assassinationAttempt} and it can also be seen by the visualization being a barplot rather than a histogram.
\item The variable concerning the sex of the president was skipped, as there is nothing to present when all US presidents have been male so far.
\item The report flags the variable \code{ethnicity} to have suspiciously few observations in one category, "African American". 
\item A few presidents were found to have odd values in the variable describing the duration of their presidencies: Some had very short (outlier) presidencies of less than two years and one was registered to have an \textit{infinite} presidency. 
\item The variable concerning age at inauguration was coded as a character variable, but consists exclusively of numbers and takes a lot of different values and therefore, it was flagged as a potentially misclassified numeric variable.
\item It seems as if one or more presidents have complex numbers as their favorite numbers. As \code{complex} is not a supported variable type in \code{dataMaid} and no strategy for handling this class was provided in the \code{treatXasY} argument, the variable is simply flagged as non-supported.
\end{itemize}
A lot of these mistakes are easily fixable, and we will do so below. However, some of them require more delicate knowledge of the subject matter. For instance, \code{ethnicity} is very reasonably marked as a potentially problematic variable as it includes only a single observation of "African American". However, a human reading this report will know that this does \textit{not} reflect a mistake in the data, but rather a peculiarity in the real world, and as such, it should not be cleaned out.

@@

and we find that only a single variable in \code{toyData}, \code{spotifysong} (for which all
observations have the value \code{"Irrelevant"}), is problem-free when using the default checks.
